---
title: "Programming Exercise 8: Anomaly Detection and Recommender Systems, Part II"
author: "Min Z"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_notebook: default
  html_document: default
---


## 2 Recommender Systems

In this part of the exercise, you will implement the collaborative filtering learning algorithm and apply it to a dataset of movie ratings.

### 2.1 Movie ratings dataset

This dataset consists of ratings on a scale of 1 to 5. The dataset has nu = 943 users, and nm = 1682 movies. 

```{r}
# remove all list
rm(list = ls())
```

```{r}
library(R.matlab)
ex8_movies<-readMat("ex8_movies.mat")
Y<-ex8_movies$Y
R<-ex8_movies$R
```

Y is a 1682x943 matrix, containing ratings (1-5) of 1682 movies on  943 users

R is a 1682x943 matrix, where R(i,j) = 1 if and only if user j gave a rating to movie i

From the matrix, we can compute statistics like average rating.

Average rating for movie 1 (Toy Story), where has been given a rate by user: 
```{r}
mean(Y[1,as.logical(R[1,])])
```

We can "visualize" the ratings matrix by:
```{r}
image(Y,xlab="movies",ylab="users",col=topo.colors(12))
```


### 2.2 Collaborative filtering learning algorithm
Now, you will start implementing the collaborative filtering learning algo- rithm. You will start by implementing the cost function (without regulariza- tion).

#### 2.2.1 Collaborative filtering cost function

#### 2.2.2 Collaborative filtering gradient

### 2.3 Learning movie recommendations
#### 2.3.1 Recommendations